What We Eval in the Shadows

### 2021-08-10 paper/main.tex
tired

Contrary to the Javascript \eval, which does not specify the environment in
which to evaluate, the second argument of \eval in R, \c{envir}, is dedicated to
that. This argument determines what is visible to the computation started by
\eval and the potential reach of its side-effects. Environments used by \eval
can be classified into the following four kinds:

\begin{compactitem}[---]
\item \emph{Function:} environments for the local variables of some function
  currently active on the call stack. Obtained by calling \c{parent.frame()} or
  \c{sys.frame()}.
\item \emph{Synthetic:} environments built from data structures such as lists,
  dataframes, or constructed explicitly with \c{new.env}, \c{list2env}, or
  \c{as.environment}. It also includes the empty environment.
\item \emph{Global:}  environments in which scripts or interactive commands are
  evaluated. \item \emph{Package:}  environments of a loaded library.
\end{compactitem}

As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in the
scope of a function with global as a distant second. This means that most
variable reads and most side-effects operate on local variables. But for which
function? From the point of view of the caller of \eval,
Table~\ref{tab:funoffset} identifies that function by its offset on the call
stack. Thus 0 is the direct caller, 1 is its parent, and so on. The data
suggests that in 81\% of cases, \eval accesses the caller -- this means the
variables of the function where \eval textually occurs are read and written to.
It is noteworthy that 1.5\% of sites evaluate three frames or above. This means
that, in general, modular reasoning is impossible in R. To understand what any
given code snippet may do requires fully understanding all the functions that
may be called, transitively, from that snippet. The actions of \eval happen at a
distance. Finally, note that any site may see several kinds, but in
\packageNbOneCategoryEnvirSitePercent of the cases they have a single kind. It
means that a compiler could predict the \emph{kind} of environment after the
first invocation.

@@

Global \evals are likely split between intentional and accidental one. Direct
references to the top-level, using \c{globalenv()} or \c{.GlobalEnv}, are rare;
they occur in only \packageNbExplicitGlobalSites sites. Thus we suspect most
uses of global are accidental. They arise from properties of our corpus, most
scripts are top-level code snippets. Thus, global is often the caller of \eval
or close to it. The reason we make this point is that values stored in the
global environment are visible to all functions and are not reclaimed by the
garbage collector. So accidental uses may pollute that namespace.

Each synthetic environment has parent that is specified when calling
\c{new.env}. When \c{envir} is a list or a data frame, \eval uses its third
argument (\c{enclos}). The parent is used to search for variables not found in
its child (side-effects stay in the child). Table~\ref{tab:newenvs} shows parent
kinds for synthetic environments. Most of them are functions, then come global
and package.

\paragraph{Discussion.}
The data presented in this section is what one could expect. For {\it
  functions}, they represent the main user case for \eval. Thus \eval is really
extending the behavior of that function by executing a dynamically selected
piece of code in the function's scope. Typically, variables are read (mostly),
written (less), but there are also some cases where new variables are injected
or existing variables are deleted. Usually, this happens in the current
function, but also in frames arbitrarily far up the call stack. One data point
we lack is how the target environment was obtained. The expected case is that
the body of a promise was modified and the resulting expression was evaluated in
the same environment the promise originated from. Likely less frequent are cases
where \eval is provided the results of programmatically selecting some call
stack. For {\it synthetics}, their relatively high frequency correlates to use
cases where one wants to evaluate an expression in either a restricted
environment or to use a data structure as an environment. For {\it global}, its
relatively high frequency is likely an artifact of how the code of our corpus is
is run. For {\it packages}, only \packageNbPackageNamespaceEnvSites sites have
this kind. This is probably for the best as mutating the bindings of
a loaded package is improper and R tries to make it difficult.

@@

Where does the expression passed to \eval come from? There are various means of
creating that expression, which are associated with particular use cases. During tracing, at each function call, we check if the return value is an expression or a symbol, or if the function is a function of interest. If it is, we record it and also look at whether the arguments were recorded before if it is not an expression-building function. It makes it possible to build a direct acyclic graph of the functions that build expressions or combine expressions during tracing to create a direct acyclic graph of the functions creating the expression passed to \eval. The following example yields the provenance graph with five nodes on Figure~\ref{fig:provgraph} for expression \c{e} passed to \eval.

\begin{lstlisting}
e <- parse(text = "a ; b")
e[[2]] <- quote(c)
eval(e)
\end{lstlisting}

\begin{figure}[h]
\centering
\includegraphics[width=8cm]{provenance-graph}
\caption{The provenance DAG for the code example above. Assignments and complex assignments are intermediate nodes, whereas expression builders are leaves of the graph.} \label{fig:provgraph}
\end{figure}

Expression \c{e} has two origins, \c{parse} and \c{quote}, which corresponds to the leaves of the graph. We also extract from the graph a representative origin, by traversing the graph from the root, following the left-hand side member of the assignments until a leaf. It represents the provenance that is further modified to yield the expression passed to \eval. In our example, the representative provenance is \c{parse}.

% Wrap tables do not play well with lists (they warn about it in the documentation actually!)
\begin{wraptable}{r}{5cm}\small\centering
	\begin{tabular}{lcc}
		\toprule
		\bf Provenance & \bf \#sites & \bf \%sites \\
		\midrule
		\packageProvenanceNamea & \packageNbProvenanceSitesa & \packagePercentProvenanceSitesa \\
		\packageProvenanceNameb & \packageNbProvenanceSitesb & \packagePercentProvenanceSitesb \\
		\packageProvenanceNamec & \packageNbProvenanceSitesc & \packagePercentProvenanceSitesc \\
		%\packageProvenanceNamed & \packageNbProvenanceSitesd & %\packagePercentProvenanceSitesd \\
		\packageProvenanceNamee & \packageNbProvenanceSitese & \packagePercentProvenanceSitese \\
		\packageProvenanceNamef & \packageNbProvenanceSitesf & \packagePercentProvenanceSitesf \\
		\packageProvenanceNameg & \packageNbProvenanceSitesg & \packagePercentProvenanceSitesg \\
		\packageProvenanceNameh & \packageNbProvenanceSitesh & \packagePercentProvenanceSitesh \\
		\packageProvenanceNamei & \packageNbProvenanceSitesi & \packagePercentProvenanceSitesi \\
		%\packageProvenanceNamej & \packageNbProvenanceSitesj & %\packagePercentProvenanceSitesj \\%[[
		%\packageProvenanceNamek & \packageNbProvenanceSitesk & %\packagePercentProvenanceSitesk \\%$
		\packageProvenanceNamel & \packageNbProvenanceSitesl & \packagePercentProvenanceSitesl \\
		%\packageProvenanceNamem & \packageNbProvenanceSitesm & %\packagePercentProvenanceSitesm \\ %as.vector
		\packageProvenanceNamen & \packageNbProvenanceSitesn & \packagePercentProvenanceSitesn \\
		\bottomrule
	\end{tabular}
	\caption{First provenances in number of sites.} \label{tab:detailed_provenances}
\end{wraptable}

Table~\ref{tab:detailed_provenances} shows the first provenances in number of sites. In \packagePercentProvenanceSitesd\% of the sites, the expression passed to \eval is not a language expression or a symbol but a floating point, or a string for instance. In that case, we did not track the provenance. \c{.External} invokes a C function. For instance, in the following code, \c{.External(C\_deriv, expr, namevec, function.arg, tag, hessian)} calls a C function called \c{C\_deriv} to derive an expression.
 Due to technical reasons, there is some imprecision in the
results. In particular, we misclassify some sites but these cases happen rarely. For instance,  \packagePercentProvenanceSitesj\% of the sites have provenance \c{[[}, a subsetting operator, as we lose the provenance of the subsetted expression, often when that expression is given attributes. Manual inspection of numerous examples suggests that errors are rare.

 We further classify them into four high-level categories:

\begin{compactitem}[---]
\item {\it Reflection:} This group corresponds to uses of \c{match.call} to
	reflectively capture the expression that invoked the current function. Function arguments are passed as promises, and \c{substitute} is used to retrieve the source
	expression associated with the promise. \c{formals} access the argument list of a function.
\item {\it String:} Expressions created from strings by invoking \c{parse},
	\c{str2expression}, or \c{str2lang}.
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{expression} functions, or by building a call with \c{call} or \c{as.call}.
\item {\it Environment: }   A symbol is created with \c{as.name} or \c{as.symbol}, typically to find the value of a symbol in a non-local environment.
\item {\it External: } a call to a C function. % Put an example
\end{compactitem}

%\NOTE{I only see 70 sites with formulas (using the class of the expression) so it does not appear here.}

%\NOTE{I can also put the absolute number of provenances, \ie if an expression has multiple provenances, we do not keep only the representative but also the other provenances. Also, if one site sometimes gets match.call and sometimes substitute, it is not going to be counted in either of them currently.}

Table~\ref{tab:provenance} summarizes the expression provenance in our corpus. Note that some sites can be invoked several times with different provenances. In that case, we have counted them several times in the table.

\begin{wraptable}[8]{r}{5cm}\small\centering
\begin{tabular}{rrr} \toprule
\bf Origin  & \bf \#sites & \bf \%sites \\\midrule
Reflection &  \packageNbReflectionSites & \packageReflectionSitesPercent\\
String & \packageNbStringSites & \packageStringSitesPercent \\
Constructed & \packageNbConstructedSites & \packageConstructedSitesPercent \\
Environment & \packageNbSymbolSites & \packageSymbolSitesPercent \\
External & \packageNbExternalSites & \packageExternalSitesPercent \\\bottomrule
\end{tabular}
\caption{Provenance}\label{tab:provenance}
\end{wraptable}

Strings could correlate with dynamic code loading. This is what the base
functions \c{source} and \c{sys.source} do. We observed few calls
(\packageNbParseFromFileSites in total) that consume the result of calling
\c{parse} on a file. Most of the calls build strings programmatically. We also
identified one function \c{invokeRestartInteractively} that prompts the user for
input, parses it, and passes it to \eval.

\paragraph{Discussion}
The origin data suggests that reflection is the first use case for \eval, \ie using \c{match.call}, \c{substitute}, or \c{match.call} to access and transform the arguments of a function when calling the function. Constructing expressions from strings is also
a common use case. The constructed category shows that most
\evals comes from code that was processed by the compiler, and may be slightly
modified by the programmer before invoking \eval. Both constructed and
reflection categories roughly correspond to meta-programming. Some of these use
cases could likely be replaced by macros if the R designers could be convinced
to overcome their distaste for those. Another use of \eval is to directly build symbols to evaluate them in another environment than the default one.

@@

\Eval may perform side-effects. From a compiler's perspective, we care about
effects that can be observed --- \ie variable definitions, updates, and removals
that are visible after a call finishes. Knowing in which environments these
side-effects happen, can help us determine how much of the compiler knowledge
about the program will be invalidated. Our analysis records information about
every environment. From the recorded data, we discard side-effects coming from
unit testing frameworks.\footnote{In the corpus, we have \c{RUnit, testthat,
  tinytest} and \c{unitizer} testing frameworks.} They run their tests via
\eval, and thus everything becomes a side-effect.

From our \packageNbrunsRnd programs, we capture \SEAllRnd side-effects from
\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
side-effects caused by the R virtual machine implementation which are not
related to the user code. For example, the \c{.Random.seed} variable, which
contains the state of the random number generator, is saved and restored from
and to the global environment every time  a statistical routine is called.
%
Removing those leaves us with \SEUserRnd side effects from \SEUserCallsRnd
\eval calls in \SEUserSites sites. These sites are part of \SEUserFunctions R
functions (\SEUserFunctionsToAllRatio of all functions doing \eval) in
\SEUserPackages packages. From these functions, \SEFunsNighty are responsible
for 90\% of side effects. Half of the side-effects come just from three
functions: \c{plyr::allocate\_column} (allocates space for a new data frame
column), \c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally, \SESitesBothEnvirRatio does both.

>>>>

Contrary to JavaScript, in R the environment of \eval can be specifies by its
\c{env} argument. This gives users control over what is visible to the
computation started by \eval and the potential reach of its side-effects. We
distinguish the following kinds of environments:

\begin{itemize}[---]
\item \emph{Function:} environments for the local variables of some function
  currently active on the call stack. Obtained by calling \c{parent.frame()} or
  \c{sys.frame()}.
\item \emph{Synthetic:} environments built from data structures such as lists,
  data.frames, or constructed explicitly with \c{new.env}, \c{list2env}, or
  \c{as.environment} or the empty environment.
\item \emph{Global:} environments in which scripts or interactive commands are
  evaluated.
\item \emph{Package:} environments of libraries.
\end{itemize}

Table~\ref{tab:highlevelenvironments} shows that most calls evaluate in the
scope of a function, global is a distant second. This means that most reads and
writes act on local variables. But of which function? Table~\ref{tab:funoffset}
gives the offset of that function on the call stack: 0 is function that called
\eval, 1 is that function's caller, and so on. In 81\% of cases, \eval is
resolved the caller -- the variables of the function where the call to \eval
textually occurs are read and written. About 1.5\% of sites evaluate their
argument three frames up the call stack, or above. Modular reasoning is thus
impossible in R. Since the actions of \eval happen at a distance, understanding
any given code snippet requires knowing which functions may be called
transitively from that snippet. In \packageNbOneCategoryEnvirSitePercent of
sites, only kind of environment is observed.

@@

Each synthetic environment has parent that is specified when calling
\c{new.env}. When \c{envir} is a list or a data frame, \eval uses its third
argument (\c{enclos}). The parent is used to search for variables not found in
its child (side-effects stay in the child). Table~\ref{tab:newenvs} shows parent
kinds for synthetic environments. Most of them are functions.

Global \evals split between intentional and accidental one. Direct references to
the top-level, using \c{globalenv()} or \c{.GlobalEnv}, are rare; they occur in
only \packageNbExplicitGlobalSites sites. Thus we suspect most uses of global
are accidental. They arise from the fact that our corpus, consists of many code
snippet that are run at top-level, global is thus the caller of \eval. This
noteworthy because writes to global variables are visible to all functions and
are not reclaimed by the garbage collector. So accidental uses may pollute that
namespace. Table~\ref{tab:polyenvir} gives the number of kinds seen at a given
call site.

\paragraph{Discussion.}
The data presented here is not surprising. The main use of \eval is to provide a
customizable extension mechanism for the behavior of functions. They are a way
to parameterize the function with any behavior that can be expressed in R. The
behavior of \eval within its enclosing function is to read and write variables,
mostly read, and very rarely delete variables. There are also some cases where
new variables are injected in a function. Usually, this in the direct caller,
but can sometimes take effect several frames up the call stack. There is
something brittle about code relying on the position of a caller, many
refactorings may break code that does that.

One bit of information that we lack is how the environment was obtained. Our
expectation is that the expression was extracted from a promise using
\c{substitute}, perhaps modified, and then evaluated with the environment coming
from the same promise. We believe the case where \eval is provided the results
of programmatically selecting some call stack to be less frequent.

Synthetic environments are relatively frequent. Common uses-case include
evaluation of an expression in a sandbox, or using a data structure as
environment, for example one could evaluate a method of an object and use and
environment to hold the object's fields.

We have already explained the popularity of global environments, as for package
environments, they are used in only \packageNbPackageNamespaceEnvSites sites.
This is probably for the best as mutating the bindings of a loaded package is
frowned upon and R tries to discourage it.

@@

Where does the expression passed to \eval come from? There are various means of
creating an expression, often these means correlate with a particular use case.
Our analysis records the values returned by some functions of interest as well
as their arguments. We build a direct acyclic graph that track the provenance of
the values that are given to \eval. The following example yields the provenance
graph with five nodes on Figure~\ref{fig:provgraph} for expression \c{e}.

\begin{lstlisting}
 > e <- parse(text = "a ; b")
 > e[[2]] <- quote(c)
 > eval(e)
\end{lstlisting}

\begin{figure}[h]\centering
\includegraphics[width=5cm]{provenance-graph}
\caption{Provenance} \label{fig:provgraph}
\end{figure}

The provenance of \c{e} includes both \c{parse} and \c{quote}. We pick an origin
by traversing the graph from the root, following the left-hand side member of
assignments until a leaf. It represents the provenance that is further modified
to yield the expression passed to \eval. In our example, the origin is \c{parse}.

% Wrap tables do not play well with lists (they warn about it in the documentation actually!)
\begin{wraptable}{r}{5cm}\small\centering
\begin{tabular}{lcc}\toprule
\bf Provenance & \bf \#sites & \bf \%sites \\\midrule
		\packageProvenanceNamea & \packageNbProvenanceSitesa & \packagePercentProvenanceSitesa \\
		\packageProvenanceNameb & \packageNbProvenanceSitesb & \packagePercentProvenanceSitesb \\
		\packageProvenanceNamec & \packageNbProvenanceSitesc & \packagePercentProvenanceSitesc \\
		\packageProvenanceNamee & \packageNbProvenanceSitese & \packagePercentProvenanceSitese \\
		\packageProvenanceNamef & \packageNbProvenanceSitesf & \packagePercentProvenanceSitesf \\
		\packageProvenanceNameg & \packageNbProvenanceSitesg & \packagePercentProvenanceSitesg \\
		\packageProvenanceNameh & \packageNbProvenanceSitesh & \packagePercentProvenanceSitesh \\
		\packageProvenanceNamei & \packageNbProvenanceSitesi & \packagePercentProvenanceSitesi \\
		\packageProvenanceNamel & \packageNbProvenanceSitesl & \packagePercentProvenanceSitesl \\
		\packageProvenanceNamen & \packageNbProvenanceSitesn & \packagePercentProvenanceSitesn \\\bottomrule
\end{tabular}
\caption{Origin} \label{tab:detailed_provenances}
\medskip
\begin{tabular}{rrr} \toprule
\bf Origin  & \bf \#sites & \bf \%sites \\\midrule
Reflection &  \packageNbReflectionSites & \packageReflectionSitesPercent\\
String & \packageNbStringSites & \packageStringSitesPercent \\
Constructed & \packageNbConstructedSites & \packageConstructedSitesPercent \\
Environment & \packageNbSymbolSites & \packageSymbolSitesPercent \\
External & \packageNbExternalSites & \packageExternalSitesPercent \\\bottomrule
\end{tabular}
\caption{Provenance}\label{tab:provenance}
\end{wraptable}

Table~\ref{tab:detailed_provenances} shows frequencies of provenance. For
technical reasons, we misclassify a small number of sites. Manual inspection of
numerous examples suggests that errors are rare. We further classify them into
high-level categories:

\begin{itemize}[---]
\item {\it Reflection:} use of \c{match.call} to reflectively capture the
  expression that invoked this function, arguments are promises, and
  \c{substitute} is used to retrieve their source.
\item {\it String:} create from strings by invoking \c{parse},
  \c{str2expression}, or \c{str2lang}.
\item {\it Constructed:} invoked \c{quote} or \c{expression}, or by built with
  \c{call} or \c{as.call}.
\item {\it Environment: } created with \c{as.name} or \c{as.symbol}, typically
  to read a non-local environment.
\item {\it External: }  call to a C function. % Put an example
\end{itemize}

Table~\ref{tab:provenance} summarizes provenance. Some sites are invoked with
different provenances, they are counted multiple times. Strings correlate with
dynamic code loading as done by \c{source} and \c{sys.source}. Few calls
(\packageNbParseFromFileSites in total) consume the result of calling \c{parse}
on a file. Most build strings programmatically. We identified one function
\c{invokeRestartInteractively} that prompts the user for input and \evals it.

\paragraph{Discussion}
The main use case for \eval involves using \c{match.call}, \c{substitute}, or
\c{match.call} to access and transform arguments of a function. Constructing
expressions from strings is also a common use case. The constructed category
shows that most \evals comes from existing code that was modified by the
programmer before invoking \eval. Both constructed and reflection categories
roughly correspond to meta-programming. Some use cases could likely be replaced
by macros if the R designers could be convinced to overcome their distaste for
those. Another use of \eval is to directly build symbols to evaluate them in
another environment.

@@

\Eval may perform side-effects. We care about effects that are visible after a
call finishes, \ie variable definitions, updates, and removals. Knowing in which
environments these side-effects happen, can help determine the impact of \eval
on our ability to reason about the code. Our analysis records information about
every environment. From the recorded data, we discard side-effects coming from
unit testing frameworks as they use \eval to run all their tests and thus
dominate the side-effect data.\footnote{The corpus has \c{RUnit, testthat,
  tinytest} and \c{unitizer} testing frameworks.}

We record \SEAllRnd side-effects from \SEAllCallsRnd calls and \SEAllSites
sites. The challenge is to remove accidental side-effects caused by the R
virtual machine implementation and not user code. For example, the
\c{.Random.seed} variable is saved and restored from and to the global
environment every time a statistical routine is called. Filtering accidental
side-effects leaves us with \SEUserRnd writes from \SEUserCallsRnd calls (in
\SEUserSites sites and \SEUserFunctions functions). \SEFunsNighty functions are
responsible for 90\% of side effects. Half of those come from just three
functions: \c{plyr::allocate\_column} (allocates space for a new data frame
column), \c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the sites (\SESitesInEnvirRatio) update
the environment specified by the \c{env} parameter.
